---
title: "A goodness-of-fit and calibration toolbox for logistic regression"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A goodness-of-fit and calibration toolbox for logistic regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
library(ebrahim.gof)
```

## Why a toolbox

Before the inferences or predictions of a logistic regression model can be trusted, its
*goodness of fit* has to be checked. Many tests exist for this, but in practice they are
**scattered across different packages with different interfaces**, base R ships only the
Hosmer--Lemeshow idea, and most of the classical tests **lose power on sparse data** --- the
common situation where continuous covariates give almost every observation its own covariate
pattern.

`ebrahim.gof` addresses this in one place. It:

* **introduces the author's own sparse-data tests** --- the omnibus Ebrahim--Farrington (EF)
  test, the *directed* EF / **EDGE** test that targets smooth calibration-shape departures, and a
  Cauchy-combination ensemble; and
* **aggregates a wide range of classical and modern tests** (Hosmer--Lemeshow, McCullagh,
  Osius--Rojek, le Cessie--van Houwelingen, Stute--Zhu, the binary-adaptive `BAGofT` test, and
  the `givitiR` calibration test) behind **one call**, `run.all.gof()`, so they can be compared
  head to head. Each aggregated test is obtained from its own package and is credited to its
  authors; they are provided for comparison, not claimed as original here.

## The example data

The bundled `gof_demo` dataset was generated from a model whose true linear predictor is
**quadratic in age**. If we fit a model that treats age *linearly*, it is therefore subtly
misspecified through a smooth, low-dimensional calibration distortion --- exactly the misfit a
*directed* test is built to detect and an *omnibus* test tends to miss.

```{r data}
data("gof_demo", package = "ebrahim.gof")
str(gof_demo)

fit <- glm(outcome ~ age + bmi + sex + treatment,
           data = gof_demo, family = binomial)
```

## One call: the whole battery

`run.all.gof()` runs the battery and returns a tidy `gof_battery` table. Here we use the fast
subset (`include_slow = FALSE`) and `install = "no"` so nothing is installed or resampled while
the vignette builds; tests whose optional package is not present are simply skipped.

```{r battery}
set.seed(1)
battery <- run.all.gof(fit, include_slow = FALSE, install = "no")
battery
```

Read the table by *family*. The **global omnibus** statistics (Pearson, deviance) spread their
power over hundreds of covariate patterns and tend to **miss** this smooth misfit. The
**directed** tests --- `DEF.poly2`, `DEF.poly3` (EDGE), `DEF.stukel`, and Stukel's score test ---
concentrate their few degrees of freedom on the calibration-curve shape and **detect** it, as do
the EF test and the ensemble. That contrast is the whole point of a directed test.

## Sparse or grouped? Let the package show you both

Three of the battery's tests are *pattern-sensitive*: `Pearson`, `Deviance`, and `McCullagh`
have a reference distribution that depends on whether they are computed per
observation (the "sparse" form) or per distinct *covariate pattern* (the "grouped" form). With
all-continuous covariates, as in `gof_demo`, every observation is its own pattern (800 patterns
for 800 observations), so the two forms **coincide**; the package recognises this degenerate
case and omits the redundant `(grouped)` rows, showing only the sparse form in the table above.
(This is also the regime where the per-observation Pearson and deviance statistics lose their
chi-squared reference entirely.) The paired `(grouped)` rows appear only when covariate
patterns actually repeat --- as they do next.

The bundled `gof_demo_grouped` dataset makes the distinction real. It uses the *same*
data-generating process, but `age` is recorded in 10-year bins and `bmi` as integers, so the 800
observations share only 328 covariate patterns. The battery detects this automatically and the
two forms now disagree --- on this data the sparse deviance says "fine" (p about 0.17) while the
grouped deviance rejects (p about 0.04), with the degrees of freedom dropping from 795 to 323:

```{r grouped}
data("gof_demo_grouped", package = "ebrahim.gof")
fit_g <- glm(outcome ~ age + bmi + sex + treatment,
             data = gof_demo_grouped, family = binomial)
set.seed(1)
run.all.gof(fit_g, include_slow = FALSE, install = "no")
```

Same model formula, same test family, opposite verdicts --- so handling the grouping wrong
silently changes conclusions. The package's answer is to never make that choice for you
invisibly: both forms are printed side by side whenever they differ, and in truly sparse data
(where no grouped "correction" can rescue Pearson or the deviance) the partition and directed
families below are the tools that remain valid.

## The directed tests on their own

Each test is also callable directly. The EDGE test projects the grouped standardized residuals
onto a small basis of smooth calibration shapes:

```{r directed}
edge.gof(fit)                 # directed EDGE test (poly3 basis)
def.gof(fit, basis = "poly2") # a lower-order directed basis
ef.gof(fit)                   # the omnibus EF test
```

The Cauchy-combination ensemble pools the directed bases into a single p-value:

```{r ensemble}
def.ensemble.gof(fit)
```

## Beyond the battery

`run.all.gof()` selects among the classical tests and the directed ones. Several of the
package's procedures sit outside it, because they need something the battery does not have:
a penalty, a frozen set of weights, or a scorer of your own. They are called directly on the
same fitted model.

### Penalised (ridge) fits

Under a penalty the classical references are not merely weak but **invalid**: shrinkage
biases the fitted probabilities, so the grouped residuals acquire a non-centrality and the
usual chi-squared reference is wrong. Two functions apply the same correction and differ
only in what they refer it to.

```{r}
set.seed(4)
n <- 300; p <- 8
X <- matrix(rnorm(n * p), n, p)
y <- rbinom(n, 1, plogis(0.2 + X %*% c(0.9, -0.6, 0.4, rep(0, p - 3))))

# closed form: one fit, no resampling, and no Monte Carlo error in the p-value
calm.gof(X, y, lambda = 40)
```

`shrink.gof()` computes the same statistics and calibrates them by prepivoting instead,
which costs `B` penalised refits. Prefer `calm.gof()` unless you need `penalize` to choose
which columns are shrunk, or `p >= n`, which it refuses. Note the penalty scale: `lambda`
here is `n` times a **glmnet** lambda, and `calm.gof()` takes a `lambda_scale` argument
while `shrink.gof()` does not.

`?calm.gof` states where the reference is validated. It is not validated for strongly
unbalanced outcomes once `p/n` is an appreciable fraction, and the function warns when you
call it there.

### Frozen-weight and pretrained tests

`legoft()` combines eleven statistics with weights fixed offline and shipped frozen, so two
analysts obtain the same p-value; `legoft.localize()` then says which domain of evidence
carries the misfit, with familywise error control. `deepgof1()` is a pretrained convolutional
statistic whose level comes from your own parametric bootstrap rather than from the network.
None of these is retrained when you call it.

### Building your own

`gof.features()` turns a fitted model into a feature vector of test statistics, and
`deploy.gof()` refers a scorer of yours to a parametric bootstrap under the fitted model.

## Going further

* **The full battery.** Set `include_slow = TRUE` to add the resampling/refit-based tests
  (le Cessie--van Houwelingen, Stute--Zhu, `BAGofT`, GAM-smooth tests) and
  `calibration_plot = TRUE` for the `givitiR` calibration test and belt. These need their
  respective optional packages; `gof_install_suggests()` installs them.
* **Speed.** The EDGE test is calibrated in closed form (no bootstrap, no refit), so it costs a
  few milliseconds even at large `n`. An optional GPU backend for the heavier O(n^2)/O(n^3)
  aggregated tests (le Cessie, projection) is available in the package's replication materials.
* **Choosing a test.** For the structured misfit that dominates practice --- a wrong link or an
  omitted smooth term --- prefer a directed test (EDGE) over an omnibus one; keep an omnibus test
  in reserve for rough, high-frequency departures.

## References

Farrington CP (1996). "On Assessing Goodness of Fit of Generalized Linear Models to Sparse Data."
*Journal of the Royal Statistical Society B*, **58**(2), 349--360.

Hosmer DW, Lemeshow S (1980). "Goodness of Fit Tests for the Multiple Logistic Regression Model."
*Communications in Statistics -- Theory and Methods*, **9**(10), 1043--1069.
